---
title: "Basic Simulation with the nphsim R Package"
author: "Keaven Anderson and Yang Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We demonstrate the basic simulation capabilities for time-to-event clinical trials that are implemented in the `nphsim` package.
This includes:

- Simulating enrollment with piecewise constant enrollment rates
- Simulating times-to-events with a piecewise exponential distribution
- Simulating a clinical trial (fixed design) by combining the above capabilities
- Performing various statistical tests on a simulated clinical trial
- Approximating power with simulation for a group sequential derived using the `gsDesign` package


## Simulating Enrollment

We use piecewise constant enrollment and piecewise exponential failure rates as methods to approximate arbitrary patterns of enrollment and dropouts. We specify an example non-proportional hazards scenario.

We approximate an arbitrary enrollment pattern by assuming constant enrollment rates within fixed time periods.
The length of enrollment periods prior to the final stable enrollment (2, 4, respectively) are specified here: 

```{r}
enrollIntervals <- c(2, 4)
```

Thus, we assume enrollment is constant in months 1-2, 3-6, and from month 7 onward.
Next we specify relative rates of enrollment in these time periods: 

```{r}
enrollRates <- c(1, 2, 4)
```

Let's say we wish to enroll 200 patients, starting with 10 patients per month.
We generate random enrollment times as follows:

```{r}
library(nphsim)
enrollTimes <- rpwexp(n = 300, rate = 10 * enrollRates, intervals = enrollIntervals, cumulative = TRUE)
```

When 'cumulative = TRUE` is input to `rpwexp`, it is designed to simulate enrollment with rates specified in `rate`. 
We plot the enrollment over time:

```{r, fig.cap = "Plot of simulated enrollment."}
library(ggplot2)
qplot(x = c(0, enrollTimes), y = 0:length(enrollTimes), geom="step", 
      ylab="Number enrolled", xlab="Time") +
      scale_x_continuous(breaks=c(0,2,6,10))
```

You can see the enrollment rate increases from the first 2 months to months 3-6 and again in the period after month 6.
The last patient in this simulation is enrolled at `max(enrollTimes) = ` `r round(max(enrollTimes),1)` months.

## Generating Piecewise Exponential Failure Times

We can also use the `rpwexp` function to generate piecewise exponential failure times.
We have a different set of intervals to define the periods of time for which different failure rates apply:

```{r}
# Failure rates for piecewise exponential time periods
failRates <- c(.8, .4, .2, .1)
# Interval durations before final stable rate
# Note that length is 1 fewer than for failRates
failIntervals <- c(0, 3, 6)
```

Now we generate the piecewise exponential failure times corresponding to the patient enrollment previously simulated, again using `rpwexp`.
We do not need to specify the default value `cumulative = FALSE` which distinguishes this from the call generating enrollment times.

```{r}
n <- length(enrollTimes)
y <- rpwexp(n = n, rate = failRates, intervals = failIntervals)
```

The longest duration among these is `max(y) = ` `r max(y)`.
Although we will eventually be interested in generating censored survival times, this is not done here; there is a value for event time each simulated patient. 
We plot the duration of time until an event for all patients ordered by decreasing time which reproduces the shape of the Kaplan-Meier curve for the points; the `theme` function removes the y-axis as well as its labels and tick-marks.
Note that at random there are some patients with very long simulated times.

```{r, fig.cap = "Plot of sorted simulated failure times without censoring."}
id <- 1:n
# sort patient time-to-event and create a 0 starting point for each patient
dta <- data.frame(N=factor(c(id,id)),Time=c(array(0,n), sort(y, decreasing = TRUE)))
ggplot(dta, aes(x = Time, y = N, grp = N)) + geom_line() +
       xlab("Time-to-event") + ylab("Patients ordered by time-to-event") +
  scale_x_continuous(breaks=(0:4)*6) + 
  theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
```



# Simulating a 2-Arm Trial Instance with Non-Proportional Hazards

We now proceed to simulating a clinical trial with the control group rate specified above and with an experimental group having piecewise exponential rates generated according to a non-proportional hazards assumption.
The parameters here appear somewhat different than above in `rpwexp` examples as they follow variable names used in the `gsDesign` R package. [SHOULD WE CHANGE THIS???]

```{r}
# Hazard ratio corresponding to control group event rates above
# and corresponding failure rates for experimental group
hr <- c(1, .7, .5, .3)
lambdaE <- hr * failRates
# Sample size of control and experimental arms
ssC <- n
ssE <- n
# We assume dropouts at a rate of .05, increasing to .1 at end
# dropout rates specified for same intervals as failure rates
eta <- c(.1,.1,.1,.3)
# Simulate a single trial instance
# We add an arbitrary enrollment interval for indefinite enrollment duration
trial <- nphsim(nsim = 1, lambdaC = failRates, lambdaE = lambdaE,
                intervals = failIntervals, ssC = ssC, ssE = ssE,
                gamma = 2 * enrollRates, R = c(enrollIntervals, 1000),
                eta = eta, etaE = eta)
# show a few lines
head(trial$simd, n=5)
```
In the above, we see the simulation ID in `sim`, treatment group in `treatment`, time of study entry in `enterT`, duration followed for survival in `survival` and censoring value (0 = event, 1 = censored) in `cnsr`. We examine censoring by treatment group:

```{r}
table(trial$simd$treatment, trial$simd$cnsr)
```

Summarizing times of events relative to start of enrollment, we have

```{r}
summary(trial$simd$enterT+trial$simd$survival)
```

We compare the distributions of times to events by computing Kaplan-Meier curves.
The curves appear to separate more as time goes on and the hazard ratio becomes more extreme.

```{r}
library(survival)
plot(with(trial$simd, survfit(Surv(survival, 1- cnsr) ~ treatment)))
```

# Performing an analysis

Now suppose we wish to test after 300 events (NOT YET WORKING).

```{r, eval=FALSE}
testOut <- simtest(x = trial, anaD=300)
testOut$result
```

